---
title: "Assignment 1.2"
author: "Marcos Cunchillos Just"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(MASS)
library(ggplot2)
library(knitr)
```

## Exercise 1

The R **datasets** package provides a series of datasets to work with, such as `iris`, `cars` or `Titanic`. You should:

- Find a statistical summary of the variables of the **iris** and **Orange** datasets.

```{r ex1-summary}
# Summarize iris
data(iris)
head(iris)
summary(iris)

# Summarize Orange
data(Orange)
head(Orange)
summary(Orange)
```

- Generate a table of absolute and relative frequencies. Do all the generated tables make sense to you?

```{r ex1-frequencies-iris}
# Absolute frequencies of the iris variables
kable(table(iris$Sepal.Length))
kable(table(iris$Sepal.Width))
kable(table(iris$Petal.Width))
kable(table(iris$Petal.Length))
kable(table(iris$Species))

# Relative frequencies of the iris variables
kable(prop.table(table(iris$Sepal.Length)))
kable(prop.table(table(iris$Sepal.Width)))
kable(prop.table(table(iris$Petal.Length)))
kable(prop.table(table(iris$Petal.Width)))
kable(prop.table(table(iris$Species)))
```

Not all of them. The table for `Species` is the only one that is really informative: 50 flowers per species, so a relative frequency of 0.333 each, which just reflects the balanced design of the sampling. The other four variables are continuous measurements, so almost every value appears once or twice and the table is as long as the dataset. A histogram or the `summary()` is the right tool there, not a frequency table.

- Generate a table of absolute frequencies for each of the variables of the **Orange** dataset. Do all the generated tables make sense to you?

```{r ex1-frequencies-orange}
# Absolute frequencies of the Orange variables
kable(table(Orange$Tree))
kable(table(Orange$age))
kable(table(Orange$circumference))
```

Here it is a bit mixed. `Tree` (5 trees, 7 measurements each) and `age` (7 measurement days, 5 trees each) make sense because they are identifiers of the design, not real measurements. `circumference` is continuous, so its table is mostly counts of 1 and tells us nothing.

- Generate a two-way table between the variables `Tree` and `age` of **Orange**.

```{r ex1-two-way-table}
table(Orange$Tree, Orange$age)
```

Every cell is 1: each tree was measured once on each of the 7 days, so the design is completely balanced and there are no missing measurements.

## Exercise 2

Copy and run the following code for the two vectors:

```{r ex2-vectors}
vect1 <- c(1,2,1,2,1,2,1,2,1,2,1,1,1,1,2,2,1,1,2,1)
vect2 <- c(1,1,2,2,2,1,2,1,1,2,1,2,1,1,1,2,1,1,1,1)
```

**a)** Using `vect1`, create a new vector called `low_weight` that is a factor with two levels. The labels correspond to 1 = "Low weight" and 2 = "Normal weight".

```{r ex2a-low-weight-factor}
low_weight <- factor(vect1, levels = c(1, 2), labels = c("Low weight", "Normal weight"))
```

**b)** Using `vect2`, create a new vector called `smoker` that is a factor with two levels. The labels correspond to 1 = "Smoker" and 2 = "Non-smoker".

```{r ex2b-smoker-factor}
smoker <- factor(vect2, levels = c(1, 2), labels = c("Smoker", "Non-smoker"))
```

**c)** Create a contingency table with the two previous variables, named `contingency_table`.

```{r ex2c-contingency-table}
contingency_table <- table(low_weight, smoker)
contingency_table
```

**d)** Examine the relationship between the two variables using the chi-squared test.

```{r ex2d-chi-squared, warning=TRUE}
# warning = TRUE on purpose: with so few observations R warns that the
# chi-squared approximation may be incorrect, which is why Fisher's test follows
chisq.test(contingency_table)
```

**e)** Also check the result of Fisher's exact test.

```{r ex2e-fisher}
fisher.test(contingency_table)
```

The chi-squared test gives a p-value of 1 and R warns that the approximation may be incorrect. That is expected: with only 20 observations some expected counts are 2.8, below the usual minimum of 5, so the chi-squared result should not be trusted here. Fisher's exact test does not have that limitation and gives p = 1 as well (odds ratio of about 1.2). In both cases we cannot reject independence: in this sample there is no evidence of a relationship between low weight and smoking, although with so few observations the test has very little power.

## Exercise 3

**a)** Load the **airquality** dataset from the **datasets** package and generate the following plots:

- A scatter plot of the variable `Ozone` in blue, using hash symbols (#) instead of dots.

```{r ex3a-scatter-ozone}
data("airquality")
plot(airquality$Ozone, col = "blue", pch = "#")
```

- A red boxplot of the variable `Temp`, with the title "Temperature (in degrees Fahrenheit)".

```{r ex3a-boxplot-temp}
boxplot(airquality$Temp, col = "red", main = "Temperature (in degrees Fahrenheit)")
```

**b)** Load the **airmiles** dataset from the **datasets** package and generate the following plots:

- A line plot of the `airmiles` series, titled "Commercial flight passenger data (in thousands)", in blue (`cadetblue2`), with the x-axis label "Thousands of passengers".

```{r ex3b-line-airmiles}
data("airmiles")
plot(airmiles,
     main = "Commercial flight passenger data (in thousands)",
     xlab = "Thousands of passengers",
     col  = "cadetblue2")
```

- A histogram of the `airmiles` series in brown (`chocolate2`).

```{r ex3b-histogram-airmiles}
hist(airmiles, col = "chocolate2")
```

**c)** Display the four plots together in a single image.

```{r ex3c-combined-plots}
par(mfrow = c(2, 2))

plot(airquality$Ozone, col = "blue", pch = "#")
boxplot(airquality$Temp, col = "red", main = "Temperature (in degrees Fahrenheit)")
plot(airmiles,
     main = "Commercial flight passenger data (in thousands)",
     xlab = "Thousands of passengers",
     col  = "cadetblue2")
hist(airmiles, col = "chocolate2")

# Restore the default single-plot layout
par(mfrow = c(1, 1))
```

## Exercise 4

Try to reproduce example 7 of this lab with data simulated by you. They do not need to be similar, but you must be able to produce different statistical plots with the `plot()` command.

```{r ex4-simulated-plots}
set.seed(999)
x1 <- rnorm(1000)         # simulate a variable x1
y1 <- x1 + rnorm(1000)    # simulate y1 as x1 plus random noise

# Scatter plot with the least-squares line (y1 as a function of x1)
plot(x1, y1, main = "Scatter plot")
abline(lm(y1 ~ x1), col = "cadetblue2")

# Density curves of both variables
plot(density(x1), main = "Densidad de x1 e y1", col = "aquamarine")
lines(density(y1), col = "darkred")
```

## Exercise 5

To practice creating plots with **ggplot2**, we will create six plots with different characteristics, using datasets from the packages used previously.

**a)** Using the **airquality** data from the **datasets** package, create a simple scatter plot with the fitted regression line in red (`darkred`).

```{r ex5a-scatter-regression-line}
ggplot(airquality, aes(Solar.R, Temp)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", colour = "darkred")
```

**b)** Using the **airquality** data from the **datasets** package, create a scatter plot of the variables `Solar.R` and `Temp` with their respective axis labels. The colours and the shape of the points must depend on the variable `Month` converted to a factor.

```{r ex5b-scatter-colour-shape}
ggplot(airquality, aes(Solar.R, Temp, colour = factor(Month), shape = factor(Month))) +
  geom_point() +
  labs(x = "Solar radiation", y = "Temperature")
```

**c)** Using the **birthwt** data from the **MASS** package, show the distribution of the variable `age` with a black border and a blue interior in the bars.

```{r ex5c-age-distribution}
ggplot(birthwt, aes(age)) +
  geom_bar(fill = "steelblue", colour = "black")
```

**d)** Using the **birthwt** data from the **MASS** package, show the distribution of the variable `age`, split into two plots (smoker / non-smoker) by the variable `smoke`.

```{r ex5d-age-by-smoking}
birthwt$smoke_f <- factor(birthwt$smoke,
                          levels = c(0, 1),
                          labels = c("Non-smoker", "Smoker"))

ggplot(birthwt, aes(x = age)) +
  geom_histogram(fill = "steelblue", colour = "darkblue") +
  facet_wrap(~ smoke_f)
```

## Exercise 6

Create a plot with data taken from R's **datasets** package and save it as an image (.jpg) named `migrafic1` and also as a PDF document named `migrafic2`.

```{r ex6-export-graphs}
# Build the plot once and reuse it
p_age <- ggplot(birthwt, aes(age)) +
  geom_bar(fill = "steelblue", colour = "black")

# Export as JPG
jpeg("migrafic1.jpg")
print(p_age)    # ggplot objects must be printed explicitly inside a device
dev.off()       # dev.off() must be CALLED, with parentheses, to write the file

# Export as PDF
pdf("migrafic2.pdf")
print(p_age)
dev.off()
```

## Exercise 7

To practice simple linear regression we will use the **Orange** dataset, found in the **tidyverse** library, which has information on three variables (tree, age in days since the tree was planted, and trunk circumference in centimetres) from 35 orange trees.

- We want to know what circumference a tree will have six hundred days after planting. Create a linear model for this and practise linear regression step by step, following the steps seen in this lab.

### Step 1: access the data

```{r ex7-step1-data}
data("Orange")
head(Orange)
```

### Step 2: summarise the variables to analyse

```{r ex7-step2-summary}
str(Orange)
summary(Orange)
```

### Step 3: scatter plots of the variables, two at a time

```{r ex7-step3-pairs}
pairs(Orange)
```

### Step 4: Pearson correlation coefficient

```{r ex7-step4-correlation}
cor(Orange$circumference, Orange$age)
```

### Step 5: fit the regression model

If we find a relationship between the two variables, we fit the linear model.

```{r ex7-step5-model}
model <- lm(circumference ~ age, data = Orange)
summary(model)
```

### Step 6: scatter plot and fitted least-squares line

```{r ex7-step6-plot}
ggplot(Orange, aes(x = age, y = circumference)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "red")
```

### Step 7: check the results

We are interested in the estimated parameters of the model and in the goodness of fit of the line, given by the coefficient of determination (R-squared). We can also look at the p-value, which in inference gives the probability of obtaining a value equal to or more extreme than the observed one, assuming a null hypothesis is true. A result is statistically significant if its p-value is less than or equal to 0.05.

```{r ex7-step7-residuals}
res <- rstandard(model)
predicted_values <- fitted(model)

plot(predicted_values, res,
     xlab = "Fitted values", ylab = "Standardized residuals")
```

### Step 8: validate the model

```{r ex7-step8-qqplot}
qqnorm(res)
qqline(res)
```

The model explains most of the variability: the correlation between age and circumference is about 0.91 and R-squared is about 0.83, with a clearly significant slope (p < 0.001). On average, a tree gains roughly 0.107 cm of circumference per day. The residuals against the fitted values show a slight funnel shape (the spread grows with the fitted value), so the constant-variance assumption is only approximately met. The Q-Q plot follows the line reasonably well except for a few points in the tails, so normality is acceptable. The model is good enough as a description, but the prediction should be taken as an approximation.

### Step 9: point prediction

If required, we can obtain a point prediction from our model with `predict()` and the new data.

```{r ex7-step9-prediction}
pred_age <- data.frame(age = 600)
predict(model, pred_age)
```

A tree 600 days after planting is expected to have a trunk circumference of about 81.5 cm. 600 days is inside the observed range of ages (118 to 1582 days), so this is an interpolation and not an extrapolation, which makes it more reliable.

## Exercise 8

Load the **PlantGrowth** dataset from R's **datasets** package. This file contains the results of an experiment to compare mean yields, measured as the dry weight of the plants (`weight`), obtained under a control and two different treatment conditions (`group` factor).

- Do you think there are differences between treatments?

### Step 1: load the data

```{r ex8-step1-data}
data("PlantGrowth")
head(PlantGrowth)
```

### Step 2: summarise the data

```{r ex8-step2-summary}
summary(PlantGrowth)
```

### Step 3: look for significant outliers

```{r ex8-step3-boxplot}
# palette.colors() starts with black, so the first colour is skipped
boxplot(weight ~ group, data = PlantGrowth,
        xlab = "Group", ylab = "Dry weight",
        col = unname(palette.colors(4))[2:4])
```

### Step 4: check approximate normality in each group

Check that the variable has an "approximately" normal distribution for each category of the independent variable.

```{r ex8-step4-shapiro}
by(PlantGrowth$weight, PlantGrowth$group, shapiro.test)
```

### Step 5: check homogeneity of variance

Check that the variances of the response are equal in every group.

```{r ex8-step5-bartlett}
bartlett.test(PlantGrowth$weight, PlantGrowth$group)
```

### Step 6: ANOVA

Check whether there are differences between the groups with an ANOVA test.

```{r ex8-step6-anova}
model_anov <- aov(weight ~ group, data = PlantGrowth)
summary(model_anov)
```

### Step 7: post hoc test

Check which groups differ, using a post hoc test such as Tukey's multiple comparison test.

```{r ex8-step7-tukey}
TukeyHSD(model_anov)
```

The ANOVA is significant (F = 4.85, p = 0.016), so at least one group mean differs from the others. The assumptions hold: Shapiro-Wilk is non-significant in the three groups and Bartlett gives p = 0.24. Tukey's test shows that the only significant pair is treatment 2 versus treatment 1 (difference of about 0.87, p = 0.012). Neither treatment differs significantly from the control, since their intervals include 0. So yes, there are differences between treatments, but only between the two treatments themselves: treatment 2 gives the highest dry weight and treatment 1 the lowest.

## Exercise 9

Using biomedical or health-related data of your choice, imported into R:

- Perform a complete statistical summary of the dataset and explain the results.

```{r ex9-import-summary}
# Import the data (Framingham heart study)
url <- "https://raw.githubusercontent.com/matackett/sta210/master/data/framingham.csv"
heart_data <- read.csv(url)
str(heart_data)
summary(heart_data)
```

The dataset has 4,240 participants and 16 variables from the Framingham cohort. The mean age is about 50 years (range 32 to 70) and 43% are men. About half of the participants smoke (mean of 9 cigarettes per day overall), 31% have hypertension and only 3% have diabetes, so the diabetic group is small. The mean total cholesterol is 237 mg/dL, systolic pressure 132 mmHg and BMI 25.8, all slightly above the healthy range on average. Glucose is very right-skewed (median 78, maximum 394) because of the diabetic patients. There are missing values, mostly in `glucose` (388), `education` (105), `BPMeds` (53) and `totChol` (50), which `lm()` and `cor()` handle by dropping those rows. About 15% of participants develop coronary heart disease within 10 years.

- Create two basic plots with the variables, explain their meaning and save them as images (jpeg or bmp).

```{r ex9-base-plots}
# Scatter plot: BMI vs heart rate
plot(heart_data$BMI, heart_data$heartRate,
     xlab = "BMI",
     ylab = "Heart rate",
     col  = "red")

# Boxplot: glucose by diabetes status
heart_data$diabetes_f <- factor(heart_data$diabetes,
                                levels = c(0, 1),
                                labels = c("No diabetes", "Diabetes"))

boxplot(glucose ~ diabetes_f, data = heart_data,
        xlab = "Diabetes status", ylab = "Glucose (mg/dL)",
        col = c("lightgreen", "tomato"))
```

The scatter plot of BMI against heart rate shows a very diffuse cloud with no clear pattern. The correlation is practically zero (around 0.07), so BMI is not a useful predictor of heart rate here. The boxplot shows that glucose is much higher and much more dispersed in people with diabetes (median around 145 mg/dL) than in people without it (median around 78 mg/dL), as expected from the definition of the disease.

- Create two plots with the `ggplot()` command, explain their meaning and save them as images (jpeg or bmp).

```{r ex9-ggplots}
heart_data$currentSmoker_f <- factor(heart_data$currentSmoker,
                                     levels = c(0, 1),
                                     labels = c("Non-smoker", "Smoker"))

# Distribution of total cholesterol by smoking status
ggplot(heart_data, aes(x = totChol)) +
  geom_histogram(fill = "steelblue", colour = "darkblue", binwidth = 10) +
  facet_wrap(~ currentSmoker_f) +
  labs(x = "Total cholesterol", y = "Count")

# Total cholesterol by smoking status
ggplot(heart_data, aes(x = currentSmoker_f, y = totChol)) +
  geom_boxplot(colour = "black") +
  labs(x = "Smoking habit", y = "Total cholesterol") +
  theme(legend.position = "none")
```

The histograms show that total cholesterol has a similar, slightly right-skewed distribution in smokers and non-smokers, with a few extreme values above 400 mg/dL. The boxplot confirms it: the medians are almost identical (about 232 vs 236 mg/dL), so smoking status does not seem to be related to total cholesterol in this sample.

- Fit a linear regression between two of its variables step by step and comment on the results.

```{r ex9-regression}
# Correlation between systolic blood pressure and age
cor(heart_data$sysBP, heart_data$age)

# Linear model: systolic blood pressure as a function of age
model_heart <- lm(sysBP ~ age, data = heart_data)
summary(model_heart)

# Scatter plot with the fitted line
ggplot(heart_data, aes(x = age, y = sysBP)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "red")

# Model validation: standardized residuals vs fitted values
res_heart <- rstandard(model_heart)
predicted_heart <- fitted(model_heart)

plot(predicted_heart, res_heart,
     xlab = "Fitted values", ylab = "Standardized residuals")

# Normality of the residuals
qqnorm(res_heart)
qqline(res_heart)
```

There is a moderate positive correlation between age and systolic blood pressure (r about 0.39). The model estimates about 82 mmHg at age 0 (not meaningful on its own) and an increase of about 1 mmHg per year of age. The slope is highly significant (p < 0.001), but R-squared is only 0.16, so age explains about 16% of the variability in blood pressure and many other factors are missing. In the residuals-versus-fitted plot the variance increases for higher fitted values, and the Q-Q plot departs from the line in the upper tail (right-skewed residuals), so the normality and homoscedasticity assumptions are not fully satisfied. A log transformation of `sysBP` or a multiple model with more predictors (BMI, hypertension, sex) would be the natural next step.
