---
title: "Assignment 1.2"
author: "Marcos Cunchillos Just"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(MASS)
library(ggplot2)
library(knitr)
```

## Exercise 1

The R **datasets** package provides a series of datasets to work with, such as `iris`, `cars` or `Titanic`. You should:

- Find a statistical summary of the variables of the **iris** and **Orange** datasets.

```{r ex1-summary}
# Summarize iris
data(iris)
head(iris)
summary(iris)

# Summarize Orange
data(Orange)
head(Orange)
summary(Orange)
```

- Generate a table of absolute and relative frequencies. Do all the generated tables make sense to you?

```{r ex1-frequencies-iris}
# Absolute frequencies of the iris variables
kable(table(iris$Sepal.Length))
kable(table(iris$Sepal.Width))
kable(table(iris$Petal.Width))
kable(table(iris$Petal.Length))
kable(table(iris$Species))

# Relative frequencies of the iris variables
kable(prop.table(table(iris$Sepal.Length)))
kable(prop.table(table(iris$Sepal.Width)))
kable(prop.table(table(iris$Petal.Length)))
kable(prop.table(table(iris$Petal.Width)))
kable(prop.table(table(iris$Species)))
```

- Generate a table of absolute frequencies for each of the variables of the **Orange** dataset. Do all the generated tables make sense to you?

```{r ex1-frequencies-orange}
# Absolute frequencies of the Orange variables
kable(table(Orange$Tree))
kable(table(Orange$age))
kable(table(Orange$circumference))
```

- Generate a two-way table between the variables `Tree` and `age` of **Orange**.

```{r ex1-two-way-table}
table(Orange$Tree, Orange$age)
```

## Exercise 2

Copy and run the following code for the two vectors:

```{r ex2-vectors}
vect1 <- c(1,2,1,2,1,2,1,2,1,2,1,1,1,1,2,2,1,1,2,1)
vect2 <- c(1,1,2,2,2,1,2,1,1,2,1,2,1,1,1,2,1,1,1,1)
```

**a)** Using `vect1`, create a new vector called `low_weight` that is a factor with two levels. The labels correspond to 1 = "Low weight" and 2 = "Normal weight".

```{r ex2a-low-weight-factor}
low_weight <- factor(vect1, levels = c(1, 2), labels = c("Low weight", "Normal weight"))
```

**b)** Using `vect2`, create a new vector called `smoker` that is a factor with two levels. The labels correspond to 1 = "Smoker" and 2 = "Non-smoker".

```{r ex2b-smoker-factor}
smoker <- factor(vect2, levels = c(1, 2), labels = c("Smoker", "Non-smoker"))
```

**c)** Create a contingency table with the two previous variables, named `contingency_table`.

```{r ex2c-contingency-table}
contingency_table <- table(low_weight, smoker)
contingency_table
```

**d)** Examine the relationship between the two variables using the chi-squared test.

```{r ex2d-chi-squared, warning=TRUE}
# warning = TRUE on purpose: with so few observations R warns that the
# chi-squared approximation may be incorrect, which is why Fisher's test follows
chisq.test(contingency_table)
```

**e)** Also check the result of Fisher's exact test.

```{r ex2e-fisher}
fisher.test(contingency_table)
```

## Exercise 3

**a)** Load the **airquality** dataset from the **datasets** package and generate the following plots:

- A scatter plot of the variable `Ozone` in blue, using hash symbols (#) instead of dots.

```{r ex3a-scatter-ozone}
data("airquality")
plot(airquality$Ozone, col = "blue", pch = "#")
```

- A red boxplot of the variable `Temp`, with the title "Temperature".

```{r ex3a-boxplot-temp}
boxplot(airquality$Temp, col = "red", main = "Temperature")
```

**b)** Load the **airmiles** dataset from the **datasets** package and generate the following plots:

- A line plot of the `airmiles` series, titled "Miles)", in blue (`cadetblue2`), with the x-axis label "Passengers".

```{r ex3b-line-airmiles}
data("airmiles")
plot(airmiles,
     main = "Miles)",
     xlab = "Passengers",
     col  = "cadetblue2")
```

- A histogram of the `airmiles` series in brown (`chocolate2`).

```{r ex3b-histogram-airmiles}
hist(airmiles, col = "chocolate2")
```

**c)** Display the four plots together in a single image.

```{r ex3c-combined-plots}
par(mfrow = c(2, 2))

plot(airquality$Ozone, col = "blue", pch = "#")
boxplot(airquality$Temp, col = "red", main = "Temperature")
plot(airmiles,
     main = "Miles",
     xlab = "Passengers",
     col  = "cadetblue2")
hist(airmiles, col = "chocolate2")

# Restore the default single-plot layout
par(mfrow = c(1, 1))
```

## Exercise 4

Try to reproduce example 7 of this lab with data simulated by you. They do not need to be similar, but you must be able to produce different statistical plots with the `plot()` command.

```{r ex4-simulated-plots}
set.seed(999)
x1 <- rnorm(1000)         # simulate a variable x1
y1 <- x1 + rnorm(1000)    # simulate y1 as x1 plus random noise

# Scatter plot with the least-squares line (y1 as a function of x1)
plot(x1, y1, main = "Scatter plot")
abline(lm(y1 ~ x1), col = "cadetblue2")

# Density curves of both variables
plot(density(x1), main = "Densidad de x1 e y1", col = "aquamarine")
lines(density(y1), col = "darkred")
```

## Exercise 5

To practice creating plots with **ggplot2**, we will create six plots with different characteristics, using datasets from the packages used previously.

**a)** Using the **airquality** data from the **datasets** package, create a simple scatter plot with the fitted regression line in red (`darkred`).

```{r ex5a-scatter-regression-line}
ggplot(airquality, aes(Solar.R, Temp)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", colour = "darkred")
```

**b)** Using the **airquality** data from the **datasets** package, create a scatter plot of the variables `Solar.R` and `Temp` with their respective axis labels. The colours and the shape of the points must depend on the variable `Month` converted to a factor.

```{r ex5b-scatter-colour-shape}
ggplot(airquality, aes(Solar.R, Temp, colour = factor(Month), shape = factor(Month))) +
  geom_point() +
  labs(x = "Solar radiation", y = "Temperature")
```

**c)** Using the **birthwt** data from the **MASS** package, show the distribution of the variable `age` with a black border and a blue interior in the bars.

```{r ex5c-age-distribution}
ggplot(birthwt, aes(age)) +
  geom_bar(fill = "steelblue", colour = "black")
```

**d)** Using the **birthwt** data from the **MASS** package, show the distribution of the variable `age`, split into two plots (smoker / non-smoker) by the variable `smoke`.

```{r ex5d-age-by-smoking}
birthwt$smoke_f <- factor(birthwt$smoke,
                          levels = c(0, 1),
                          labels = c("Non-smoker", "Smoker"))

ggplot(birthwt, aes(x = age)) +
  geom_histogram(fill = "steelblue", colour = "darkblue") +
  facet_wrap(~ smoke_f)
```

## Exercise 6

Create a plot with data taken from R's **datasets** package and save it as an image (.jpg) named `migrafic1` and also as a PDF document named `migrafic2`.

```{r ex6-export-graphs}
# Build the plot once and reuse it
p_age <- ggplot(birthwt, aes(age)) +
  geom_bar(fill = "steelblue", colour = "black")

# Export as JPG
jpeg("migrafic1.jpg")
print(p_age)    # ggplot objects must be printed explicitly inside a device
dev.off()       # dev.off() must be CALLED, with parentheses, to write the file

# Export as PDF
pdf("migrafic2.pdf")
print(p_age)
dev.off()
```

## Exercise 7

To practice simple linear regression we will use the **Orange** dataset, found in the **tidyverse** library, which has information on three variables (tree, age in days since the tree was planted, and trunk circumference in centimetres) from 35 orange trees.

- We want to know what circumference a tree will have six hundred days after planting. Create a linear model for this and practise linear regression step by step, following the steps seen in this lab.

### Step 1: access the data

```{r ex7-step1-data}
data("Orange")
head(Orange)
```

### Step 2: summarise the variables to analyse

```{r ex7-step2-summary}
str(Orange)
summary(Orange)
```

### Step 3: scatter plots of the variables, two at a time

```{r ex7-step3-pairs}
pairs(Orange)
```

### Step 4: Pearson correlation coefficient

```{r ex7-step4-correlation}
cor(Orange$circumference, Orange$age)
```

### Step 5: fit the regression model

If we find a relationship between the two variables, we fit the linear model.

```{r ex7-step5-model}
model <- lm(circumference ~ age, data = Orange)
summary(model)
```

### Step 6: scatter plot and fitted least-squares line

```{r ex7-step6-plot}
ggplot(Orange, aes(x = age, y = circumference)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "red")
```

### Step 7: check the results

We are interested in the estimated parameters of the model and in the goodness of fit of the line, given by the coefficient of determination (R-squared). We can also look at the p-value, which in inference gives the probability of obtaining a value equal to or more extreme than the observed one, assuming a null hypothesis is true. A result is statistically significant if its p-value is less than or equal to 0.05.

```{r ex7-step7-residuals}
res <- rstandard(model)
predicted_values <- fitted(model)

plot(predicted_values, res,
     xlab = "Valores ajustados", ylab = "Residuos estandarizados")
```

### Step 8: validate the model

```{r ex7-step8-qqplot}
qqnorm(res)
qqline(res)
```

### Step 9: point prediction

If required, we can obtain a point prediction from our model with `predict()` and the new data.

```{r ex7-step9-prediction}
pred_age <- data.frame(age = 600)
predict(model, pred_age)
```

## Exercise 8

Load the **PlantGrowth** dataset from R's **datasets** package. This file contains the results of an experiment to compare mean yields, measured as the dry weight of the plants (`weight`), obtained under a control and two different treatment conditions (`group` factor).

- Do you think there are differences between treatments?

### Step 1: load the data

```{r ex8-step1-data}
data("PlantGrowth")
head(PlantGrowth)
```

### Step 2: summarise the data

```{r ex8-step2-summary}
summary(PlantGrowth)
```

### Step 3: look for significant outliers

```{r ex8-step3-boxplot}
# palette.colors() starts with black, so the first colour is skipped
boxplot(weight ~ group, data = PlantGrowth,
        xlab = "Group", ylab = "Dry weight",
        col = unname(palette.colors(4))[2:4])
```

### Step 4: check approximate normality in each group

Check that the variable has an "approximately" normal distribution for each category of the independent variable.

```{r ex8-step4-shapiro}
by(PlantGrowth$weight, PlantGrowth$group, shapiro.test)
```

### Step 5: check homogeneity of variance

Check that the variances of the response are equal in every group.

```{r ex8-step5-bartlett}
bartlett.test(PlantGrowth$weight, PlantGrowth$group)
```

### Step 6: ANOVA

Check whether there are differences between the groups with an ANOVA test.

```{r ex8-step6-anova}
model_anov <- aov(weight ~ group, data = PlantGrowth)
summary(model_anov)
```

### Step 7: post hoc test

Check which groups differ, using a post hoc test such as Tukey's multiple comparison test.

```{r ex8-step7-tukey}
TukeyHSD(model_anov)
```

## Exercise 9

Using biomedical or health-related data of your choice, imported into R:

- Perform a complete statistical summary of the dataset and explain the results.

```{r ex9-import-summary}
# Import the data (Framingham heart study)
url <- "https://raw.githubusercontent.com/matackett/sta210/master/data/framingham.csv"
heart_data <- read.csv(url)
str(heart_data)
summary(heart_data)
```

- Create two basic plots with the variables, explain their meaning and save them as images (jpeg or bmp).

```{r ex9-base-plots}
# Scatter plot: BMI vs heart rate
plot(heart_data$BMI, heart_data$heartRate,
     xlab = "IMC",
     ylab = "Heart rate",
     col  = "red")

# Boxplot: glucose by diabetes status
heart_data$diabetes_f <- factor(heart_data$diabetes,
                                levels = c(0, 1),
                                labels = c("No diabetes", "Diabetes"))

boxplot(glucose ~ diabetes_f, data = heart_data,
        xlab = "Estado de diabetes", ylab = "Glucosa (mg/dL)",
        col = c("lightgreen", "tomato"))
```

- Create two plots with the `ggplot()` command, explain their meaning and save them as images (jpeg or bmp).

```{r ex9-ggplots}
heart_data$currentSmoker_f <- factor(heart_data$currentSmoker,
                                     levels = c(0, 1),
                                     labels = c("Non-smoker", "Smoker"))

# Distribution of total cholesterol by smoking status
ggplot(heart_data, aes(x = totChol)) +
  geom_histogram(fill = "steelblue", colour = "darkblue", binwidth = 10) +
  facet_wrap(~ currentSmoker_f) +
  labs(x = "Total cholesterol", y = "Frecuencia")

# Total cholesterol by smoking status
ggplot(heart_data, aes(x = currentSmoker_f, y = totChol)) +
  geom_boxplot(colour = "black") +
  labs(x = "Smoking habit", y = "Total cholesterol") +
  theme(legend.position = "none")
```

- Fit a linear regression between two of its variables step by step and comment on the results.

```{r ex9-regression}
# Correlation between systolic blood pressure and age
cor(heart_data$sysBP, heart_data$age)

# Linear model: systolic blood pressure as a function of age
model_heart <- lm(sysBP ~ age, data = heart_data)
summary(model_heart)

# Scatter plot with the fitted line
ggplot(heart_data, aes(x = age, y = sysBP)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "red")

# Model validation: standardized residuals vs fitted values
res_heart <- rstandard(model_heart)
predicted_heart <- fitted(model_heart)

plot(predicted_heart, res_heart,
     xlab = "Valores ajustados", ylab = "Residuos estandarizados")

# Normality of the residuals
qqnorm(res_heart)
qqline(res_heart)
```
